---
title: "The sf package"
output: html_notebook
---

#### 1.0 - Purpose



#### 2.0 Prereq


##### 2.1 Packages


```{r}
library(sf)
setwd("C:/WorkDocs/rGIS/vector")
```


##### 2.2 Load Data


```{r}
#Download and unzip shapefile from open data site
#Download school locations point shapefile
downloadLink <- "http://opendata.toronto.ca/gcc/school_frm_police_mtm3.zip"
download.file(url = downloadLink,
              destfile = "school_frm_police_mtm3.zip", mode = "wb")


#Download TTC subway line shapefile
downloadLink <- "http://opendata.toronto.ca/gcc/TTC_subway%20lines_mtm3.zip"
download.file(url = downloadLink,
              destfile = "TTC_subway lines_mtm3.zip", mode = "wb")


#Download ward polygon
downloadLink <- "http://opendata.toronto.ca/gcc/WARD25_OpenData_08072018_mtm3.zip"
download.file(url = downloadLink,
              destfile = "WARD25_OpenData_08072018_mtm3.zip", mode = "wb")


#Get all zip files in directory
allZip <- list.files(path = "C:/WorkDocs/rGIS/vector", pattern = "*.zip", full.names = TRUE)
plyr::ldply(.data = allZip, .fun = unzip)
```


```{r}
#Get names of unzipped files
unzipSHP <- list.files(path = "C:/WorkDocs/rGIS/vector", pattern = "*.shp", full.names = TRUE)
unzipSHP
```


```{r}
#Load shapefile using sf package

schools <- st_read("SCHOOL.shp")
subway <- st_read("TTC_SUBWAY_LINES.shp")
wards <- st_read("WARD.shp")
```


#### 3.0 Data Manipulation with sf


##### 3.1 Geometric Operations



##### 3.2 Geometric Confirmation



##### 3.3 Geometry Creation



##### 3.4 Geometry Operations


***st_crop***


***st_difference***


***st_intersection***


***st_sym_difference***


***st_snap***


***st_union***


##### 3.5 Gemetry Validation

***st_is_empty***


***st_is_simple***


***st_is_valid***


##### 3.6 Geometric Measurement


***st_area***


***st_distance***


***st_length***


##### 3.7 Misc Operations

***st_cast***


***st_coordinates***


***st_crs***


***st_join***


***st_make_grid***


***st_nearest_feature***


***st_nearest_points***


***st_transform***

